Requirements Specification

### 2025-05-15 nbs/development/architecture.qmd
docs(architecture): fix duplicate sec ref
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(development): note requirements for AnnData object mutation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

#### Validation Framework {#sec-validation-framework}

@@

>>>>

#### Validation Framework {#sec-core-validation-framework}

@@

# AnnData Integration {#sec-anndata-integration}

PyroVelocity integrates with AnnData objects following these principles:

1. **Mutation Tracking**: All methods that modify AnnData objects must track and document mutations using the `print_anndata`, `anndata_string`, and `print_string_diff` utilities.

2. **Return Values**: Methods that modify AnnData objects must return the modified object to enable method chaining and make mutations explicit.

3. **Required Parameters**: AnnData parameters should be required (not optional) when they are essential for method functionality.

4. **Shape Validation**: Array shapes should be validated using jaxtyping rather than complex shape handling code.

5. **Explicit Documentation**: Methods that modify objects must explicitly document this behavior in their docstrings.

6. **Distinction Between Model State and External Data**: While the model state follows an immutable pattern (using `ModelState` and `with_state()`), external data structures like AnnData objects can be mutated in-place for efficiency. This distinction is important because:
   - Model state must remain consistent and thread-safe
   - AnnData objects are large and copying them is expensive
   - In-place mutation of AnnData is a common pattern in the single-cell ecosystem

These principles ensure that mutations to AnnData objects are explicit, traceable, and well-documented, making the code more maintainable and reducing the risk of bugs while maintaining performance with large datasets.

### 2025-05-15 nbs/development/requirements.qmd
docs(requirements): remove unavailable bibtex key
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>
docs(development): note requirements for AnnData object mutation
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

- NumPyro documentation [@Phan2019-numpyro]
- Hoffman and Gelman (2014) [No U-Turn Sampler]

@@

>>>>

- NumPyro documentation
- Hoffman and Gelman (2014) [No U-Turn Sampler]

@@

## API Design Principles {#sec-api-design-principles}

| Principle ID | Principle Name | Description | Priority |
|--------------|----------------|-------------|----------|
| API-01 | AnnData Mutation Tracking | Methods that modify AnnData objects must track and document mutations using print_anndata, anndata_string, and print_string_diff | High |
| API-02 | AnnData Return Values | Methods that modify AnnData objects must return the modified object | High |
| API-03 | Required Parameters | Parameters that are essential for method functionality should be required, not optional | High |
| API-04 | Shape Validation | Array shapes should be validated using jaxtyping rather than complex shape handling code | High |
| API-05 | Explicit Documentation | Methods that modify objects must explicitly document this behavior in their docstrings | High |
